---
title: "Lotri Motivation"
author: "Matthew Fidler"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Lotri Motivation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This was made to allow people (like me) to specify lower triangular
matrices similar to the domain specific language implemented in
nlmixr.  Originally I had it included in `RxODE`, but thought it may
have more general applicability, so I separated it into a new
package. 

For me, specifying the matrices in this way is easier than
specifying them using R's default matrix.  For instance to fully
specify a simple 2x2 matrix, in R you specify:

```{r}
mat <- matrix(c(1, 0.5, 0.5, 1),nrow=2,ncol=2,dimnames=list(c("a", "b"), c("a", "b")))
```

With `lotri`, you simply specify:

```{r}
library(lotri)

mat <- lotri(a+b ~ c(1,
                     0.5, 1))
```

I find it more legible and easier to specify, especially if you have a
more complex matrix.  For instance with the more complex matrix:

```{r}
mat <- lotri({
    a+b ~ c(1,
            0.5, 1)
    c ~ 1
    d +e ~ c(1,
             0.5, 1)
})
```

To fully specify this in base R you would need to use:

```{r}
mat <- matrix(c(1, 0.5, 0, 0, 0,
                0.5, 1, 0, 0, 0,
                0, 0, 1, 0, 0,
                0, 0, 0, 1, 0.5,
                0, 0, 0, 0.5, 1),
              nrow=5, ncol=5,
              dimnames= list(c("a", "b", "c", "d", "e"), c("a", "b", "c", "d", "e")))
```

Of course with the excellent `Matrix` package this is a bit easier:

```{r}
library(Matrix)
mat <- matrix(c(1, 0.5, 0.5, 1),nrow=2,ncol=2,dimnames=list(c("a", "b"), c("a", "b")))
mat <- bdiag(list(mat, matrix(1), mat))
## Convert back to standard matrix
mat <- as.matrix(mat)
##
dimnames(mat) <- list(c("a", "b", "c", "d", "e"), c("a", "b", "c", "d", "e"))
```

Regardless, I think `lotri` is a bit easier to use.
