---
title: "Repeated blocks with same() in lotri"
author: "Matthew Fidler"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Repeated blocks with same() in lotri}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(lotri)
```

# What `same()` is for

NONMEM can declare an omega block and then repeat it:

```
$OMEGA BLOCK(2)
 0.1
 0.01 0.2
$OMEGA BLOCK(2) SAME
$OMEGA BLOCK(2) SAME
```

Those three blocks are not three covariance matrices.  They are *one*
estimated covariance stamped three times.  That is how inter-occasion
variability is parameterized: every occasion draws its own random
effects, but all occasions share one covariance -- and sharing one
covariance is what lets the per-occasion random effects be
**correlated**.

`same()` is the `lotri` spelling of that:

```{r}
iov <- lotri({
  iov.cl1 + iov.v1 ~ c(0.1,
                       0.01, 0.2)
  iov.cl2 + iov.v2 ~ same()
  iov.cl3 + iov.v3 ~ same()
})

iov
```

Three 2x2 blocks, one estimated 2x2:

```{r}
dim(iov)
```

# The rules

`same()` repeats the immediately preceding **block**, under new names,
and takes no arguments.  A second `same()` repeats that same original
block rather than the previous copy, the way NONMEM chains `SAME`.

It repeats the block *as declared*.  A structural zero in the
covariance does not split the block in two:

```{r}
lotri({
  a + b ~ c(1,
            0, 1)
  c1 + d1 ~ same()
})
```

A repeated block inherits the fixed flags of the block it repeats, works
under a condition, and works for a single parameter:

```{r}
lotri({
  eta.ka ~ 0.6
  iov.cl1 + iov.v1 ~ c(0.1,
                       0.01, 0.2) | occ
  iov.cl2 + iov.v2 ~ same() | occ
})
```

`same()` looks back only within one `{}` block, and only at its own
level of variability.  Each extra argument to `lotri()` is parsed by its
own call, so this has nothing to repeat:

```{r, error = TRUE}
lotri(a + b ~ c(1, 0.1, 2), c1 + d1 ~ same())
```

Write the two lines in one `lotri({})` block instead.

`same()` is refused with `rcm=TRUE` and with a `cov` function.  Both
reorder or adjust the whole matrix, which would move a repeated block
away from the block it is supposed to repeat; erroring is better than
returning a matrix whose claimed repetition is no longer true.

# How it is stored: the `condition` column

This is the part other packages need to know.

`as.data.frame()` does **not** gain a column for `same()`.  The
repetition rides in the existing `condition` column, which names the
element that is mirrored:

```{r}
df <- as.data.frame(iov)
df[, c("neta1", "neta2", "name", "est", "condition")]
```

The format is

```
<baseCondition>:same:<masterEta>                 # diagonal row
<baseCondition>:same:<masterEta1>:<masterEta2>   # covariance row
```

Two things are worth spelling out.

**The master is named, not indexed.**  `neta1`/`neta2` are renumbered
whenever parameters are added, dropped or reordered, so an index would
go stale silently.  Names are unique within one of these data frames and
survive reordering.

**Covariance rows are carried too.**  The off-diagonal row above points
at the master's off-diagonal, so the *correlation* is shared, not just
the variances.

A repeated row keeps its master's `est` and `fix`.  A consumer that
knows nothing about the suffix therefore still reads a numerically
correct matrix -- it just thinks the model has more free parameters than
it does.

# Reading the column

Do not compare `condition` directly.  A test like `condition == "id"`
misses every repeated row, and `condition != "id"` wrongly treats one as
a different level of variability.  Use these instead:

```{r}
lotriBaseCondition(df$condition)

lotriIsSame(df$condition)
```

`lotriSameMap()` gives, for each eta, the index of the eta it mirrors
(`0` when it is an ordinary or master eta).  This is the object an
estimator wants, since it says which parameters are actually free:

```{r}
lotriSameMap(df)
```

When an edit structurally changes a block -- deleting one of its
parameters, say -- the linkage no longer describes anything real.
`lotriSameBreak()` drops it, turning the copies into ordinary
independent blocks and leaving every other family alone:

```{r}
broken <- lotriSameBreak(df, "iov.v2")
broken$condition
```

# Round trips

`same()` survives every direction:

```{r}
identical(as.data.frame(as.lotri(df)), df)

identical(as.data.frame(eval(as.expression(iov))), df)

as.expression(iov)
```

and the linkage is carried across `lotriMat()`, so combining blocks does
not silently turn a repeated block back into independently estimated
parameters:

```{r}
attr(lotriMat(lotriMatInv(iov)), "lotriSame")
```

# `same()` versus `cnd(same = n)`

There are two spellings of the same NONMEM idea at different
granularities, and they compose.

- `~ same()` repeats **one block** inside a matrix, under names you
  choose.  This is `$OMEGA BLOCK(n) SAME`.
- `cnd(same = n)` repeats a **whole nesting level** `n` times, and is
  what `lotriSep()` sets up for `rxode2`'s nested simulation.

Used together, a two-parameter correlated occasion block declared with
`same()` can itself be stamped once per occasion:

```{r}
nested <- lotri(lotri({
  a + b ~ c(1,
            0.1, 2)
  c1 + d1 ~ same()
}) | occ(same = 3L))

dim(lotriMat(nested, format = "ETA[%d]", start = 1L))
```
