---
title: "Prior distributions in lotri"
author: "Matthew Fidler"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Prior distributions in lotri}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(lotri)
```

# Why priors are in `lotri`

A `lotri({})` block is the `ini({})` block of a `nlmixr2` model.  Until now
it could only say what a parameter *is* -- an estimate, optional bounds, a
`fix` flag, a label -- but not what is believed about it beforehand.

Bayesian estimation methods need that missing piece.  This adds a way to
attach a prior distribution to any parameter in the block.

`lotri` deliberately stops at parsing, validating and storing the prior.  It
does not evaluate the density and it does not generate any 'Stan' code; that
is the job of the package doing the estimation.  What `lotri` guarantees is
that by the time the estimation code sees a prior, the distribution exists,
has the right number of arguments, and does not contradict the parameter's
bounds.

# The syntax

A prior is given with `prior(name) ~ dist(...)`:

```{r}
m <- lotri({
  tka <- 0.45
  tcl <- c(0, 1, 10)

  prior(tka) ~ dnorm(0, 10)
  prior(tcl) ~ dlnorm(1, 0.5)
})

lotriEst(m)
```

Because the statement names the parameter it applies to, prior lines are
**order independent**.  These two blocks are the same:

```{r}
m1 <- lotri({
  tka <- 0.45
  prior(tka) ~ dnorm(0, 10)
})

m2 <- lotri({
  prior(tka) ~ dnorm(0, 10)
  tka <- 0.45
})

identical(m1, m2)
```

That matters in practice: you can keep all the priors together at the bottom
of a long block instead of interleaving them with the estimates.

# What a prior can be attached to

## Population parameters

```{r}
lotriEst(lotri({
  tka <- 0.45
  prior(tka) ~ dnorm(0, 10)
}))
```

## A single between subject variability term

```{r}
m <- lotri({
  eta.ka ~ 0.3
  prior(eta.ka) ~ dgamma(2, 1)
})

attr(m, "lotriPriors")
```

## A whole covariance block

Correlation and covariance matrices have their own distributions, and those
apply to the block rather than to any one element.  Give every name in the
block:

```{r}
m <- lotri({
  eta.cl + eta.v ~ c(0.1,
                     0.01, 0.2)
  prior(eta.cl, eta.v) ~ lkj_corr(2)
})

attr(m, "lotriPriors")
```

The prior is stored on the first diagonal element of the block it belongs
to, which is why only the first entry is filled in above.

The names have to be *exactly* one block.  Two unrelated diagonal elements
are two 1x1 blocks, not one 2x2 block, so this is an error:

```{r error=TRUE}
lotri({
  eta.a ~ 1
  eta.b ~ 1
  prior(eta.a, eta.b) ~ lkj_corr(2)
})
```

# Distribution names

The R name is canonical whenever R parameterizes the distribution the same
way 'Stan' does.  When there is no faithful R equivalent, the 'Stan' name is
canonical.  Both spellings are always accepted on input and are normalized to
the canonical one, so these give the same result:

```{r}
a <- lotri({ tka <- 0.45; prior(tka) ~ dnorm(0, 10) })
b <- lotri({ tka <- 0.45; prior(tka) ~ normal(0, 10) })

identical(a, b)
lotriEst(a)$prior
```

Arguments may be positional or named, in any order:

```{r}
lotriEst(lotri({
  tka <- 0.45
  prior(tka) ~ dnorm(sd=10, mean=0)
}))$prior
```

## Why `dt()` is not `student_t()`

The one place where an R name is *refused* is `dt()`.  R's `dt(x, df, ncp)`
is the standardized (or noncentral) t, while the 'Stan'
`student_t(nu, mu, sigma)` is a location-scale t.  They are different
distributions, so aliasing them would silently change the model:

```{r error=TRUE}
lotri({
  tka <- 0.45
  prior(tka) ~ dt(3)
})
```

Use `student_t()` with its own parameterization instead:

```{r}
lotriEst(lotri({
  tka <- 0.45
  prior(tka) ~ student_t(3, 0, 10)
}))$prior
```

## The supported distributions

`lotriPriorDists()` returns the whole table, including the 'Stan' name for
each distribution, which is what a package generating 'Stan' code needs:

```{r}
d <- lotriPriorDists()
nrow(d)
head(d, 10)
```

The `kind` column says what a distribution may be attached to --
`univariate` for a single parameter, `matrix` and `multivariate` for a
covariance block:

```{r}
subset(d, kind == "matrix", select=c(name, stanName, parNames))
```

# Bounds and truncated priors

The bounds are not repeated in the prior.  They already live on the
parameter, so a symmetric distribution on a parameter bounded below by zero
*is* a half distribution:

```{r}
m <- lotri({
  propSd <- c(0, 0.1)
  prior(propSd) ~ dcauchy(0, 5)
})

as.data.frame(m)[, c("name", "lower", "est", "upper", "prior")]
```

Here `propSd` has `lower = 0` and a Cauchy prior, so it is a half-Cauchy.  A
package generating 'Stan' code has everything it needs to emit the
`T[0, ]` truncation.

Because the bounds are known, `lotri` can also check that the prior does not
contradict them.  A distribution with positive support on a parameter that
allows negative values is an error:

```{r error=TRUE}
lotri({
  a <- c(-10, 1, 10)
  prior(a) ~ dlnorm(0, 1)
})
```

# What else is checked

Unknown distributions are rejected, with a suggestion when there is an
obvious near match:

```{r error=TRUE}
lotri({ a <- 1; prior(a) ~ dnorml(0, 1) })
```

So are the wrong number of arguments, and argument names that do not belong
to the distribution:

```{r error=TRUE}
lotri({ a <- 1; prior(a) ~ dnorm(0) })
```

```{r error=TRUE}
lotri({ a <- 1; prior(a) ~ dnorm(mu=0, sd=1) })
```

A matrix-valued distribution on a single parameter, a univariate one on a
block, a prior on a parameter that does not exist, and two priors on the
same parameter are all errors as well.

# Getting the priors back out

Priors round trip.  They appear in the `prior` column of the estimate data
frame and of `as.data.frame()`, and they are printed back as valid `lotri`
code:

```{r}
m <- lotri({
  tka <- 0.45
  label("Ka")
  tcl <- c(0, 1, 10)

  eta.cl + eta.v ~ c(0.1,
                     0.01, 0.2)
  eta.ka ~ 0.3

  prior(tka) ~ dnorm(0, 10)
  prior(tcl) ~ dlnorm(1, 0.5)
  prior(eta.ka) ~ dgamma(2, 1)
  prior(eta.cl, eta.v) ~ lkj_corr(2)
})

as.data.frame(m)[, c("name", "est", "condition", "prior")]
```

```{r}
as.expression(m)
```

Note that the block prior comes back as `prior(eta.cl, eta.v)`, with the
whole block recovered, not just the element it was stored on.  Since the
deparsed form is valid input, it can be fed straight back in:

```{r}
identical(as.data.frame(eval(as.expression(m))), as.data.frame(m))
```

Priors are matched to parameters **by name**, never by position, so they are
unaffected by the matrix re-ordering that `lotri` does when `rcm=TRUE` (which
is what `nlmixr2` uses):

```{r}
m <- lotri({
  a ~ 1
  b ~ c(0, 1)
  c ~ c(0.5, 0, 1)
  prior(a) ~ dgamma(1, 1)
}, rcm=TRUE)

dimnames(m)[[1]]
attr(m, "lotriPriors")
```

The matrix has been re-ordered, and the prior has followed `a` rather than
staying on the first row.

# Using this downstream

A package that generates 'Stan' code needs two things from `lotri`, and both
are available:

1. the prior itself, from the `prior` column of `as.data.frame()` (which is
   the `$iniDf` of a `nlmixr2` model) together with that row's `lower` and
   `upper` for any truncation, and
2. the 'Stan' spelling of the distribution, from `lotriPriorDists()`.

```{r}
m <- lotri({
  tka <- 0.45
  prior(tka) ~ dnorm(0, 10)
})

.df <- as.data.frame(m)
.p <- .df$prior[!is.na(.df$prior)]
.p

## map the canonical name to the Stan one
.fn <- as.character(str2lang(.p)[[1]])
lotriPriorDists()$stanName[lotriPriorDists()$name == .fn]
```

which is enough to write `target += normal_lpdf(tka | 0, 10);`.
